Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2019-02-09 pkg/disttree/inst/draft/draft.Rnw
- revised rejoinder and corresponding modifications in the manuscript - especially the discussion of more complex distributions should be checked again - also rejoinder on East Tyrol aspects should be improved
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2457 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d3a61fb932ac7bc551babc5d498793cb0b08415b

Distributional forests share some concepts and algorithmic aspects 
with other generalizations of Breiman and Cutler's random forests. Nearest 
neighbor weights are employed for aggregation in survival forests 
\citep{Hothorn+Lausen+Benner:2004},
quantile regression forests \citep{Meinshausen:2006},
transformation forests \citep{Hothorn+Zeileis:2017},
and generalized random forests for causal inferences 
\citep{Athey+Tibshirani+Wager:2017}.
These procedures aggregate over trees fitted to specific score functions 
(e.g., log rank scores in survival trees, model residuals in transformation 
or generalized forests). Distributional forests, in contrast to 
nonparametric approaches, provide a compromise between model flexibility 
and interpretability: The parameters of a problem-specific distribution 
(zero-censored Gaussian for precipitation) have a clear meaning but 
may depend on external variables in a quite general way.

@@

To assess calibration, Figure~\ref{fig:axams-qq} shows residual QQ~plots for
out-of-sample predictions (2009--2012) from the different models trained on
1985--2008. Due to the point masses at zero 100 draws from the randomized
quantile residuals \citep{Dunn+Smyth:1996} are plotted in semi-transparent
gray. Overall, the randomized quantile residuals conform quite well with the
theoretical standard normal quantile (i.e., form a straight line close to the 
diagonal), indicating that all four models are sufficiently well calibrated. 
This is also supported by the corresponding probability integral transform
(PIT, \citealp{Gneiting+Balabdaoui+Raftery:2007}) histograms for station
Axams in Supplement B \citep{Schlosser+Hothorn+Stauffer:2018b} which 
contains a more detailed explanation of residual QQ plots and PIT histograms.

@@

\subsubsection*{Distributional specifications}

Choosing an adequate distributional family is an important step for
establishing a well-fitting model. A zero-censored Gaussian distribution is
employed in this manuscript as this has been found to be an appropriate choice
for precipitation modeling in earlier literature
\citep[e.g.,][]{Stauffer+Mayr+Messner:2017}. 
To test for robustness against distributional misspecifation, 
two alternative distributional specifications have been considered
in Supplement A \citep{Schlosser+Hothorn+Stauffer:2018a}: Using the
same evaluations as in Section~\ref{sec:all}, all models are
additionally fitted for 15~meteorological stations using a zero-censored
logistic distribution in order to account for heavier tails and a two-part
Gaussian hurdle model combining a binary model for zero vs.~positive
precipitation and a separate Gaussian model, truncated at zero, for the
positive precipitation observations. Both specifications yield qualitatively 
similar results as for the zero-censored Gaussian distribution. For some stations
the two-part hurdle model leads to small improvements, however at the expense
of increased variability across stations (especially for EMOS and the 
boosted GAMLSS). Overall, the results from this manuscript are quite
robust across these distributional specifications, especially for the
distributional forests.

Moreover, one could consider a distribution including an additional parameter
for capturing skewness \citep[as in][]{Scheuerer+Hamill:2015, Baran+Nemoda:2016}. 
However, this would go beyond the mean/variance specification of the NGR
that is widely used in ensemble post-processing. Therefore, this contribution
investigates the effects of using the same distributional family with a novel
strategy for specifying dependence on covariates. Considering more flexible
parametric specifications as well as transformation model specifications that
allow to capture dependence of higher moments \citep[see e.g.,][]{Hothorn+Zeileis:2017}
are extensions of interest for future research.

\subsubsection*{Distributional complexity / Orthogonality of distribution parameters}
\hlc{Allowing for splits in each distribution parameter rather than only in the location parameter
is particularly advantageous if the parameters are orthogonal, i.e. do not depend on each other,
as changes in parameters other than the location parameter would be overlooked otherwise.
If there is a dependency among parameters it can be sufficient considering only one parameter 
in order to detect all changes, but this depends on the type of dependency.

In the case of the here applied censored Gaussian distribution a shift in the mean also influences 
the variance, hence the parameters are not orthogonal. The same holds for the 
distributions employed in Supplement A} \citep{Schlosser+Hothorn+Stauffer:2018a}. 
\hlc{Therefore, distributional trees and forests have shown to perform reasonably well also for  
distributions with non-orthogonal parameters. However, testing a broader variety of 
distributions, particularly very complex distributions such as the Dagum distribution}
(see, e.g., \cite{Klein+Kneib+Lang:2015}) 
\hlc{are required in order to confirm these findings in general.}

@@

Axams was chosen as the meteorological station for the more extensive evaluations
in Section~\ref{sec:axams} as it yields fairly typical results
and is geographically in the center of the study area and closest to
Innsbruck, the capital of Tyrol and the work place of three of the authors.
To show that qualitatively similar results are obtained for other
meteorological stations, Supplement B \citep{Schlosser+Hothorn+Stauffer:2018b}
carries out the same evaluation for 14~further stations. These
cover a wide range of geographical locations/altitudes and a mix of
different best-performing models in the single-split setting reported in
Section~\ref{sec:all}.

@@

The fitted distributional forest for July 24 and
observation station Axams (including Figure~\ref{fig:axams-dist})
is reproducible using \code{demo("RainAxams", package = "disttree")}.
This also includes fitting the other zero-censored
Gaussian models considered in this paper and generating the corresponding 
QQ~plots (Figure~\ref{fig:axams-qq}) and PIT histograms 
\citep[Supplement B]{Schlosser+Hothorn+Stauffer:2018b}.
Full replication of all results can be obtained with
\code{demo("RainTyrol", package = "disttree")}
requiring the companion \textsf{R} package \textbf{RainTyrol} 
(version~\Sexpr{packageVersion("RainTyrol")}),
also available within the \textsf{R}-Forge~project.
The results presented in Supplement A \citep{Schlosser+Hothorn+Stauffer:2018a} and 
Supplement B \citep{Schlosser+Hothorn+Stauffer:2018b} can be reproduced using
\code{demo("RainDistributions", package = "disttree")} and 
\code{demo("RainStationwise", package = "disttree")}, respectively.

@@

\section*{Supplementary material}
\subsection*{Supplement A: Different Response Distributions}
  To assess the goodness of fit of the Gaussian distribution, left-censored at zero,
  this supplement employs the same evaluations as in the main manuscript but based on
  two other distributional assumptions: A logistic distribution, left-censored at zero,
  is employed to potentially better capture heavy tails -- and a two-part hurdle model
  combining a binary model for zero vs.\ positive precipitation and a Gaussian model,
  truncated at zero, for the positive precipitation observations.

\subsection*{Supplement B: Stationwise Evaluation}
  To show that Axams is a fairly typical
  station and similar insights can be obtained for other stations as well,
  this supplement presents the same analysis as in Section~\ref{sec:axams} of the main
  manuscript for 14~further meteorological stations.

>>>>

Distributional forests share some concepts and algorithmic aspects 
with other generalizations of Breiman and Cutler's random forests. Nearest 
neighbor weights are employed for aggregation in survival forests 
\citep{Hothorn+Lausen+Benner:2004},
quantile regression forests \citep{Meinshausen:2006},
transformation forests \citep{Hothorn+Zeileis:2017},
and generalized random forests for causal inferences 
\citep{Athey+Tibshirani+Wager:2019}.
These procedures aggregate over trees fitted to specific score functions 
(e.g., log rank scores in survival trees, model residuals in transformation 
or generalized forests). Distributional forests, in contrast to 
nonparametric approaches, provide a compromise between model flexibility 
and interpretability: The parameters of a problem-specific distribution 
(zero-censored Gaussian for precipitation) have a clear meaning but 
may depend on external variables in a quite general way.

@@

To assess calibration, Figure~\ref{fig:axams-qq} shows residual QQ~plots for
out-of-sample predictions (2009--2012) from the different models trained on
1985--2008. Due to the point masses at zero 100 draws from the randomized
quantile residuals \citep{Dunn+Smyth:1996} are plotted in semi-transparent
gray. Overall, the randomized quantile residuals conform quite well with the
theoretical standard normal quantile (i.e., form a straight line close to the 
diagonal), indicating that all four models are sufficiently well calibrated. 
This is also supported by the corresponding probability integral transform
(PIT, \citealp{Gneiting+Balabdaoui+Raftery:2007}) histograms for station
Axams in Supplement~B \citep{Schlosser+Hothorn+Stauffer:2018b} which 
contains a more detailed explanation of residual QQ plots and PIT histograms.

@@

\subsubsection*{Distributional specifications for precipitation modeling}

Choosing an adequate distributional family is an important step for
establishing a well-fitting model. A zero-censored Gaussian distribution is
employed in this manuscript as this has been found to be an appropriate choice
for precipitation modeling in earlier literature
\citep[e.g.,][]{Stauffer+Mayr+Messner:2017}. 
To test for robustness against distributional misspecifation, 
two alternative distributional specifications have been considered
in Supplement~A \citep{Schlosser+Hothorn+Stauffer:2018a}: Using the
same evaluations as in Section~\ref{sec:all}, all models are
additionally fitted for 15~meteorological stations using a zero-censored
logistic distribution in order to account for heavier tails and a two-part
Gaussian hurdle model combining a binary model for zero vs.~positive
precipitation and a separate Gaussian model, truncated at zero, for the
positive precipitation observations. Both specifications yield qualitatively 
similar results as for the zero-censored Gaussian distribution. For some stations
the two-part hurdle model leads to small improvements, however at the expense
of increased variability across stations (especially for EMOS and the 
boosted GAMLSS). Overall, the results from this manuscript are quite
robust across these distributional specifications, especially for the
distributional forests.

Moreover, one could consider a distribution including an additional parameter
for capturing skewness \citep[as in][]{Scheuerer+Hamill:2015, Baran+Nemoda:2016}. 
However, this would go beyond the mean/variance specification of the NGR
that is widely used in ensemble post-processing. Therefore, this contribution
investigates the effects of using the same distributional family with a novel
strategy for specifying dependence on covariates.

\subsubsection*{More general distributional specifications}

\hlc{Beyond the task of modeling precipitation it is of interest how well distributional
forests perform in combination with other more general distributional specifications.
It has been shown previously in the literature that using a score- or gradient-based
selection of splitting variables outperforms a mean-based selection with subsequent
flexible distributional modeling: For example, both} \citet[][Figure~2]{Athey+Tibshirani+Wager:2019}
\hlc{and} \citet[][Figure~1]{Hothorn+Zeileis:2017} \hlc{demonstrated (independently) that their
respective score-based random forest algorithms outperform the mean-based quantile
regression forests of} \cite{Meinshausen:2006} \hlc{in a setup where only the variance of a normal
response variable changes across the considered covariates. However, if all distribution
parameters are closely correlated with the distribution mean the forests with different
splitting strategies all perform similarly, provided a sufficiently flexible distribution
is employed for the final predictions} \citep[see][Section~7]{Hothorn+Zeileis:2017}.

\hlc{Similarly, the score-based distributional forests introduced in this manuscript
proved to be quite robust to the different distributional specifications considered.
While all specifications focus on capturing mean-variance effects note that these
parameters are never fully orthogonal but can actually become quite closely correlated
due to the censoring (or trunctation and/or zero-inflation considered in Supplement~A).

However, exploring extensions to more flexible parametric distributions}
\citep[e.g., such as the Dagum distribution considered by][in GAMLSS-type models]{Klein+Kneib+Lang:2015}
\hlc{as well as transformation model specifications} \citep[e.g., as in][]{Hothorn+Zeileis:2017}
\hlc{are of interest for future research.}

@@

Axams was chosen as the meteorological station for the more extensive evaluations
in Section~\ref{sec:axams} as it yields fairly typical results
and is geographically in the center of the study area and closest to
Innsbruck, the capital of Tyrol and the work place of three of the authors.
To show that qualitatively similar results are obtained for other
meteorological stations, Supplement~B \citep{Schlosser+Hothorn+Stauffer:2018b}
carries out the same evaluation for 14~further stations. These
cover a wide range of geographical locations/altitudes and a mix of
different best-performing models in the single-split setting reported in
Section~\ref{sec:all}.

@@

The fitted distributional forest for July 24 and
observation station Axams (including Figure~\ref{fig:axams-dist})
is reproducible using \code{demo("RainAxams", package = "disttree")}.
This also includes fitting the other zero-censored
Gaussian models considered in this paper and generating the corresponding 
QQ~plots (Figure~\ref{fig:axams-qq}) and PIT histograms 
\citep[Supplement~B]{Schlosser+Hothorn+Stauffer:2018b}.
Full replication of all results can be obtained with
\code{demo("RainTyrol", package = "disttree")}
requiring the companion \textsf{R} package \textbf{RainTyrol} 
(version~\Sexpr{packageVersion("RainTyrol")}),
also available within the \textsf{R}-Forge~project.
The results presented in Supplement~A \citep{Schlosser+Hothorn+Stauffer:2018a} and 
Supplement~B \citep{Schlosser+Hothorn+Stauffer:2018b} can be reproduced using
\code{demo("RainDistributions", package = "disttree")} and 
\code{demo("RainStationwise", package = "disttree")}, respectively.

@@

\section*{Supplementary material}
\subsection*{Supplement~A: Different Response Distributions}
  To assess the goodness of fit of the Gaussian distribution, left-censored at zero,
  this supplement employs the same evaluations as in the main manuscript but based on
  two other distributional assumptions: A logistic distribution, left-censored at zero,
  is employed to potentially better capture heavy tails -- and a two-part hurdle model
  combining a binary model for zero vs.\ positive precipitation and a Gaussian model,
  truncated at zero, for the positive precipitation observations.

\subsection*{Supplement~B: Stationwise Evaluation}
  To show that Axams is a fairly typical
  station and similar insights can be obtained for other stations as well,
  this supplement presents the same analysis as in Section~\ref{sec:axams} of the main
  manuscript for 14~further meteorological stations.
